---
title: "Visualizing Coffee Trade data"
output: html_notebook
---

RMarkdown Notebook visualization stack presenting coffee production over time, comparing retail and grower profit margins, and measuring grower revenue against national poverty lines

```{r}
setwd("~/Desktop/Personal Projects/Coffee Data ")
#import packages 
library(RColorBrewer)
library(dplyr)
library(plotly)
library(tidyverse)
library(gridExtra)
library(ggplot2)
library(readxl)
library(waterfall)

```

### Prepare Data

```{r}
#import data files 
total_production = read.csv(file = 'total-production.csv')
retail_prices = read.csv(file = 'retail-prices.csv')
grower_compensation = read.csv(file = 'prices-paid-to-growers.csv') 
exports_number = read.csv(file = 'exports-crop-year.csv')
poverty_lines = read.csv(file = 'national-poverty-lines-vs-gdp-per-capita.csv')
coffee_consumption = read.csv(file = 'disappearance.csv')
#rename columns
colnames(retail_prices)[which(names(retail_prices) == 'retail_prices')] = 'country'
#subset columns from poverty lines
poverty_lines = data.frame(subset(poverty_lines, select = c(National.poverty.lines..Jolliffe.and.Prydz..2016..,Entity,Year)))                             
#remove NAs 
poverty_lines = na.omit(poverty_lines) 
```

### Visualizing total producion over the years
```{r}
#transpose total production df
total_production_2 = data.frame(t(total_production))
#create function to take first row as column names
header.true <- function(df) {
  names(df) <- as.character(unlist(df[1,]))
  df[-1,]
}
total_production_2 = header.true(total_production_2)
#create a column of years from 1990-2018
years = seq(1990,2018)
total_production_2 <- tibble::rownames_to_column(total_production_2, "Year")
total_production_2["Year"] = c(years)
```


```{r}
#Plot first 6 countries
#create list of plots 
plot_list = list()
#create 6 plots, add to list of plots
for (i in 2:7) {
  total_production_2[,i] = sapply(total_production_2[,i], as.numeric)
  p = ggplot(data = total_production_2,aes_string(x = "Year", y =  as.name((colnames(total_production_2[i]))))) + geom_point(stat = "identity") 
  plot_list[[i]] = p
}
#drop first plot
plot_list = plot_list[-1]
#show plot list
do.call(grid.arrange, list(grobs = plot_list, ncol=3))
```

### Compare grower vs retail profit margin  
```{r}
# Prepare data
production_costs = read_excel("Coffee Production Costs.xlsx")
production_costs = data.frame(t(production_costs))
production_costs = subset(production_costs, select = c(X7))
production_costs['country'] = rownames(production_costs)
production_costs = production_costs[-1,]
production_costs['Production Costs'] = production_costs['X7'] #assign production cost column

# Aggregate data
colnames(grower_compensation)[which(names(grower_compensation) == 'prices_paid_to_growers')] = 'country' #rename country column
costs_compensation = data.frame(inner_join(production_costs, grower_compensation)) #take overlap between porouduction cost and grower compensation 
costs_compensation = data.frame((subset(costs_compensation, select = (-c(X7)))))
A = function(x) x * 0.45359237 #From 1 pound to 1 kg 
costs_compensation['Production.Costs'] = sapply(costs_compensation['Production.Costs'], as.numeric)
x = as.numeric(colMeans(costs_compensation['Production.Costs']))
costs_compensation['Production.Costs'] = apply(costs_compensation['Production.Costs'],2, A)
```

```{r}
#function to calculate profit margin 
profit_margin = function(price, consumption, cost, production) {
  (((price * consumption) - (cost*production)) /  (price * consumption)) 
}
#function to calculate average over the years
average = function(dataframe, a){
 dataframe = dataframe[-1]
 avg_data = colMeans(dataframe, na.rm=TRUE)
 avg_data = data.frame(t(data.frame(avg_data)))
 rownames(avg_data) = a
 return (avg_data)
}
#calculate profit margin for retail side 

average_retail_price = average(retail_prices, "average_retail_price")
average_coffee_consumption = average(coffee_consumption, "average_coffee_consumption")
average_coffee_consumption[1,] = average_coffee_consumption[1,]*60*1000 #multiply by 60-kg thousand bags 
average_coffee_trade_price = average(grower_compensation, "average_coffee_trade_price")
average_exports = average(exports_number, "total_exports")
average_exports[1,] = average_exports[1,] * 60 *1000 #multiply by 60-kg thousand bags 
#average retail export-price-consumption df
retail_price_consumption = rbind(average_retail_price, average_coffee_consumption,average_coffee_trade_price, average_exports)
retail_price_consumption = data.frame(t(retail_price_consumption)) #transpose
#calculate retail revenue 
retail_price_consumption$revenue = 0 
retail_price_consumption$revenue = retail_price_consumption$average_retail_price * retail_price_consumption$average_coffee_consumption
#calculate retail costs
retail_price_consumption$cost = 0 
retail_price_consumption$cost = retail_price_consumption$average_coffee_trade_price*retail_price_consumption$total_exports
#caclculate profit margin 
retail_price_consumption$profit_margin = 0 
retail_price_consumption$profit_margin = mapply(profit_margin, retail_price_consumption$average_retail_price,retail_price_consumption$average_coffee_consumption, retail_price_consumption$average_coffee_trade_price,retail_price_consumption$total_exports)
#create years column 
years = seq(1990,2018)
retail_price_consumption <- tibble::rownames_to_column(retail_price_consumption, "Year")
retail_price_consumption["Year"] = c(years)
```


```{r}
#Calculate profit margin for grower side 
#average production of coffee
average_production = average(total_production, "average_production")
average_production[1,] = average_production[1,]*60*1000
#grower price - exports - production 
grower_price_consumption = rbind(average_coffee_trade_price,average_exports, average_production)
grower_price_consumption = data.frame(t(grower_price_consumption))
#grower production cost/lkg 
grower_price_consumption$production_costs = 1
grower_price_consumption$production_costs = grower_price_consumption$production_costs * as.numeric(colMeans(costs_compensation['Production.Costs']))
#grower revenue 
grower_price_consumption$revenue = 0 
grower_price_consumption$revenue = grower_price_consumption$average_coffee_trade_price*grower_price_consumption$total_exports
#grower costs 
grower_price_consumption$cost = 0 
grower_price_consumption$cost = grower_price_consumption$production_costs*grower_price_consumption$average_production 
#grower profit margin 
grower_price_consumption$profit_margin = 0 
grower_price_consumption$profit_margin = mapply(profit_margin, grower_price_consumption$average_coffee_trade_price,grower_price_consumption$total_exports, grower_price_consumption$production_costs,grower_price_consumption$average_production)

#create years column 
grower_price_consumption <- tibble::rownames_to_column(grower_price_consumption, "Year")
grower_price_consumption["Year"] = c(years)
```
#### Bar chart: revenue - cost - profit margin for both in each year . vertical 3 part + margin percent
```{r}
#barplot as matrix profit margin , revenue between the two 

profit_margin = data.frame(cbind(grower_price_consumption$Year, grower_price_consumption$profit_margin, retail_price_consumption$profit_margin))
colnames(profit_margin)[which(names(profit_margin) == 'X1')] = 'Year'
colnames(profit_margin)[which(names(profit_margin) == 'X2')] = 'Grower Profit Margin'
colnames(profit_margin)[which(names(profit_margin) == 'X3')] = 'Retail Profit Margin'
profit_margin = data.frame(t(profit_margin))
profit_margin = header.true(profit_margin)
barplot(height = as.matrix(profit_margin), las=2, cex.names = 0.9, col = c("chartreuse4","cadetblue3"), beside = TRUE, legend=TRUE, ylim = c(0,1.5))

revenue = data.frame(cbind(grower_price_consumption$Year, grower_price_consumption$revenue, retail_price_consumption$revenue))
colnames(revenue)[which(names(revenue) == 'X1')] = 'Year'
colnames(revenue)[which(names(revenue) == 'X2')] = 'Grower Revenue'
colnames(revenue)[which(names(revenue) == 'X3')] = 'Retail Revenue'
revenue = data.frame(t(revenue))
revenue = header.true(revenue)
opar = par(oma = c(1,0,0,8))
barplot(height = as.matrix(revenue),las=2, cex.names = 0.5, cex.axis = 0.5,  col = c("chartreuse4","cadetblue3"), beside = TRUE, yaxp=c(0, max(revenue), 5))
par(opar)
opar = par(oma = c(0,0,0,0), mar = c(0,0,0,0), new = TRUE)
legend(x = "right", legend = rownames(revenue), fill = c("chartreuse4","cadetblue3"), bty = "n", y.intersp = 2)
par(opar) # Reset par



```




### Plot farmer profits/kg of coffee to the poeverty lỉne (USD/day) for each country 
```{r}
# Prepare data
colnames(poverty_lines)[which(names(poverty_lines) == 'Entity')] = 'country'
costs_compen_country = data.frame(inner_join(costs_compensation,poverty_lines))
costs_compen_country['Profits'] =costs_compen_country$X2011 - costs_compen_country$Production.Costs
colnames(costs_compen_country)[which(names(costs_compen_country) == 'National.poverty.lines..Jolliffe.and.Prydz..2016..')] = 'Poverty Lines'
```

```{r warning=FALSE}
# Plot
fig <- plot_ly()
fig <- fig %>%
  add_trace(
    type = "indicator",
    mode = "number+gauge",
    value = as.numeric(subset(subset(costs_compen_country, country == 'Colombia'),select = c(Profits))),
    domain = list(x = c(0.25, 1), y = c(0.4, 0.6)),
    title = list(text = 'Colombia'),
    gauge = list(
      shape = "bullet",
      axis = list(range = list(NULL, 6)),
      threshold = list(
        line = list(color = "red", width = 2),
        thickness = 0.75,
        value = as.numeric(subset(costs_compen_country, country == 'Colombia')['Poverty Lines'])),
      steps = list(
        list(range = c(0,0.5), color = "white"),
        list(range = c(0.5,2), color = "white"))),
    bar = list(color = "black"))
fig <- fig %>%
  add_trace(
    type = "indicator",
    mode = "number+gauge",
    value = as.numeric(subset(subset(costs_compen_country, country == 'Guatemala'),select = c(Profits))),
    domain = list(x = c(0.25, 1), y = c(0.7, 0.9)),
    title = list(text = 'Guatemala'),
    gauge = list(
      shape = "bullet",
      axis = list(range = list(NULL, 6)),
      threshold = list(
        line = list(color = "red", width = 2),
        thickness = 0.75,
        value = as.numeric(subset(costs_compen_country, country == 'Guatemala')['Poverty Lines'])),
      steps = list(
        list(range = c(0,0.5), color = "white"),
        list(range = c(0.5,2), color = "white"))),
    bar = list(color = "black"))
fig <- fig %>%
  add_trace(
    type = "indicator",
    mode = "number+gauge",
    value = as.numeric(subset(subset(costs_compen_country, country == 'El Salvador'),select = c(Profits))),
    domain = list(x = c(0.25, 1), y = c(0.08, 0.25)),
    title = list(text = 'El Salvador'),
    gauge = list(
      shape = "bullet",
      axis = list(range = list(NULL, 6)),
      threshold = list(
        line = list(color = "red", width = 2),
        thickness = 0.75,
        value = as.numeric(subset(costs_compen_country, country == 'El Salvador')['Poverty Lines'])),
      steps = list(
        list(range = c(0,0.5), color = "white"),
        list(range = c(0.5,2), color = "white"))),
    bar = list(color = "black"))
fig

```



